---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(ensembldb)
library(ape)
library(Hmisc)
library(corrplot)
library(data.table)
library(ggrepel)
library(biomaRt)
```
First we need to rename the files in order to make the compatible with downstream steps
```{bash}
# Terminal
# Navigate to fastq folder, then delete irrelevant files
/panfs/roc/groups/7/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq
# For every given folder, create a list of filenames (example given below, but we will have several "filenames.txt" files for each sequencing batch)
ls > filenames.txt
```
```{r}
# Now download from filezilla and import into R
# Rename filename column for simplicity, add in date for reference, and give date an id "I" for "Illumina"
filenames_210723 <- read.csv("../data/fastq_filenames/filenames_210723.txt") %>%
  rename(globus_filename = 1) %>%
  mutate(seq_date = "210723") %>%
  mutate(seq_id = "I1")
filenames_211023 <- read.csv("../data/fastq_filenames/filenames_211023.txt") %>%
  rename(globus_filename = 1) %>%
  mutate(seq_date = "211023") %>%
  mutate(seq_id = "I2")
filenames_220204 <- read.csv("../data/fastq_filenames/filenames_220204.txt") %>%
  rename(globus_filename = 1) %>%
  mutate(seq_date = "220204") %>%
  mutate(seq_id = "I3")
filenames_220223 <- read.csv("../data/fastq_filenames/filenames_220223.txt") %>%
  rename(globus_filename = 1) %>%
  mutate(seq_date = "220223") %>%
  mutate(seq_id = "I4")
filenames_220307 <- read.csv("../data/fastq_filenames/filenames_220307.txt") %>%
  rename(globus_filename = 1) %>%
  mutate(seq_date = "220307") %>%
  mutate(seq_id = "I5")
```
```{r}
# The first set of filenames from 210723 are different from the others. First adjust the names to match the others
filenames_210723[,1] <- sub("_", "", filenames_210723[,1]) # remove first occurence of underscore
```
Now let's coordinate the filename objects into a list, then break down the string components
```{r}
# List of dfs
filenames_list <- list(filenames_210723, filenames_211023, filenames_220204, filenames_220223, filenames_220307)
# Duplicate globus_filename
filenames_string <- lapply(filenames_list, mutate, globus_filename2=globus_filename)
# Separate globus_filename into strings
filenames_string <- lapply(filenames_string, separate, globus_filename2, c("sample_name", "seq_number", "R1_R2", "filetype"), "_")
# Create new filename for each sample
filenames_string <- lapply(filenames_string, mutate, new_filename=paste(sample_name, seq_id, R1_R2, filetype, sep="_"))
# Extract just the original and new filenames
filenames_pairs <- lapply(filenames_string, select, globus_filename, new_filename)
```
Check each
```{r}
filenames_pairs[1]
filenames_pairs[2]
filenames_pairs[3]
filenames_pairs[4]
filenames_pairs[5]
```
Unlist, then save as individual .txt files without headers
```{r}
newfilenames_210723 <- data.frame(filenames_pairs[1])
newfilenames_211023 <- data.frame(filenames_pairs[2])
newfilenames_220204 <- data.frame(filenames_pairs[3])
newfilenames_220223 <- data.frame(filenames_pairs[4])
newfilenames_220307 <- data.frame(filenames_pairs[5])
```
Forgot we need to re-convert 210723 samples
```{r}
newfilenames_210723$globus_filename <- sub("^(.{5})", "\\1_", newfilenames_210723$globus_filename)
```
```{r}
write.table(newfilenames_210723, "../data/fastq_filenames/newfilenames_210723.txt", sep=",",  col.names=FALSE, row.names=FALSE, quote = FALSE)
write.table(newfilenames_211023, "../data/fastq_filenames/newfilenames_211023.txt", sep=",",  col.names=FALSE, row.names=FALSE, quote = FALSE)
write.table(newfilenames_220204, "../data/fastq_filenames/newfilenames_220204.txt", sep=",",  col.names=FALSE, row.names=FALSE, quote = FALSE)
write.table(newfilenames_220223, "../data/fastq_filenames/newfilenames_220223.txt", sep=",",  col.names=FALSE, row.names=FALSE, quote = FALSE)
write.table(newfilenames_220307, "../data/fastq_filenames/newfilenames_220307.txt", sep=",",  col.names=FALSE, row.names=FALSE, quote = FALSE)
```
Now back to the terminal, rename the files in individual folders. One example shown
https://unix.stackexchange.com/questions/87694/renaming-files-based-on-the-contents-of-a-text-file
```{bash}
while IFS=',' read -a files 
do 
   mv "${files[0]}" "${files[1]}" 
done < newfilenames_210723.txt
```
Now release all files to main folder, pulling out of individual folders
https://askubuntu.com/questions/146634/shell-script-to-move-all-files-from-subfolders-to-parent-folder
First move to parent folder then run"
```{bash}
find . -mindepth 2 -type f -print -exec mv {} . \;
```
There are 396 files: 198 pairs. I removed all "undetermined" samples. 
I1: 30 files from globus
I2: 70 files from globus
I3: 144 files from globus
I4: 76 files from globus
I5: 76 files from globus
Numbers add up! 
Now let's create additional files that are merged samples.
https://www.biostars.org/p/317385/
```{bash}
#!/bin/sh        
#SBATCH --time=24:00:00
#SBATCH --ntasks=8
#SBATCH --mem=32g
#SBATCH --mail-type=ALL  
#SBATCH --mail-user=sarif@umn.edu

cd /panfs/roc/groups/7/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq

for i in $(ls -1 | sed 's/[_].*$//' | uniq)

    do echo "Merging R1"

cat "$i"_I*_R1_001.fastq.gz > "$i"_ME_R1_001.fastq.gz

       echo "Merging R2"

cat "$i"_I*_R2_001.fastq.gz > "$i"_ME_R2_001.fastq.gz

done;
```
On filezilla, I moved these merged files into a separate folder (/merged). They contain not only the merged files, but the files from GMCC1 that were only run once, but still have the new "ME" string. These files are the exact same size as the original files.
The metadata Shreya sent me contains 88 files (of those samples that have been run). The merged data contains 88 pairs.
To start, I will just run the pipeline on the merged samples. If that all looks good, I'll return to running the individual un-merged files. I know there will be issues with this because for some, the library sizes are so small they error when it comes time to do edgeR.

For analysis of the pilot RNA seq data, I will be using the CHURP pipeline from MSI: https://pages.github.umn.edu/MSI-RIS/Tutorials/rnaseq_cmd/
```{bash}
module load churp/0.2.2-slurm
# Make a directory in scratch space
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Work_ME
# Make metadata
$CHURP group_template bulk_rnaseq \
    -f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/merged \
    -o /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/gmbc_ccyte_CHURP_meta.csv
```
SUCCESS

Template file: /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/gmbc_ccyte_CHURP_meta.csv

All sample groups and any additional columns have specified have been filled
with NULL. Please edit the file and write in the correct values for your
dataset. Samples with the same "Group" label will be treated as replicates
in the analysis. Samples with a "Group" value of NULL will not be used in
downstream differential expression analysis. When you have edited the file to
your liking, supply its path to the "bulk_rnaseq" pipeline with the -e
option to enable group testing.
```{bash}
$CHURP bulk_rnaseq \-e /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/gmbc_ccyte_CHURP_meta.csv \-f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/merged \-x /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/hisat2/genome \-g /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/annotation/Homo_sapiens.GRCh38.100.gtf.gz \-o /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME \-d /scratch.global/sarif/gmbc_ccyte_CHURP/Work_ME \--rmdup --ppn 6 --mem 60000 -w 24 --no-submit
```
SUCCESS

Samplesheet and pipeline script generation complete! Their paths are given
below:

Pipeline script: /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/2022-04-15.sarif.bulk_rnaseq.pipeline.sh
Samplesheet: /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/2022-04-15.sarif.bulk_rnaseq.samplesheet.txt
Sbatch array key: /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/2022-04-15.sarif.bulk_rnaseq.qsub_array.txt

Verify the information in the samplesheet, and run the pipeline script with
bash while logged into Mesabi. You will recieve email notifications of job
start/completion/error at your UMN X500 email address. If you need to submit
an error report, please contact help[at]msi.umn.edu. Please include the
samplesheet, pipeline script, and the error message with your report.
```{bash}
# Execute script
bash /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/2022-04-15.sarif.bulk_rnaseq.pipeline.sh
```
sbatch: Setting account: blekhman
sbatch: Setting account: blekhman
Output and logs will be written to /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME
Emails will be sent to sarif@umn.edu
Sbatch array to samplename key: /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/2022-04-15.sarif.bulk_rnaseq.qsub_array.txt
Single samples job array ID: 12329846
Summary job ID: 12329847

Now I am going to repeat that analysis on the un-merged files. I will likely encounter some issues with some of the file sizes being too small, but at least I will be able to single out those that are causing the issues and re-run.
```{bash}
module load churp/0.2.2-slurm
# Make a directory in scratch space
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Out
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Work
# Make metadata
$CHURP group_template bulk_rnaseq \
    -f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq \
    -o /scratch.global/sarif/gmbc_ccyte_CHURP/Out/gmbc_ccyte_CHURP_meta.csv
```
```{bash}
$CHURP bulk_rnaseq \-e /scratch.global/sarif/gmbc_ccyte_CHURP/Out/gmbc_ccyte_CHURP_meta.csv \-f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq \-x /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/hisat2/genome \-g /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/annotation/Homo_sapiens.GRCh38.100.gtf.gz \-o /scratch.global/sarif/gmbc_ccyte_CHURP/Out \-d /scratch.global/sarif/gmbc_ccyte_CHURP/Work \--rmdup --ppn 6 --mem 32000 -w 24 --no-submit
```
```{bash}
# Execute script
bash /scratch.global/sarif/gmbc_ccyte_CHURP/Out/2022-04-16.sarif.bulk_rnaseq.pipeline.sh
```

- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
Important files from this analysis:
Sample quality report (merged): 
  /scratch.global/sarif/gmbc_ccyte_CHURP/Out_ME/Bulk_RNAseq_Report.html
RNA seq technical meetrics (merged):
  /scratch.global/sarif/gmbc_ccyte_CHURP/Work_ME/allsamples/RNASeq_Metrics.txt
Subread counts (merged):
  /scratch.global/sarif/gmbc_ccyte_CHURP/Work_ME/allsamples/subread_counts.txt
Subread counts (unmerged):
  /scratch.global/sarif/gmbc_ccyte_CHURP/Work/allsamples/subread_counts.txt
- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 

After the run, I identified these files as those with a small number of mapped gene counts:
GMCC2S12_I3   0
GMCC2C1_I3    44
GMCC2S2_I3    8
GMCC2S9_I3    7
GMCC2S11_I3   56
GMCC2S13_I3   19
GMCC3S1_I3    8

I'd like to group these samples in with the QC of the other samples, so instead of starting at the .bam file level, I am just going to exclude the one holding up the pipeline (GMCC2S12_I3) and run the remainder. I predict the QC report on these samples will look pretty bad- if so, I will remove them and re-run the merged samples
```{bash}
# First, need to create a new directory that excludes that sample
# Navigate and create file
cd /panfs/roc/groups/7/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq
mkdir colonocyte_RNAseq_fastq_samples_excluded
# In bash, use an extended glob to copy over desired files and leave behind others
# https://askubuntu.com/questions/1163773/how-can-i-copy-a-directory-without-copying-a-specific-file-or-subdirectory-in-te
shopt -s extglob
cp -r colonocyte_RNAseq_fastq/!(merged) colonocyte_RNAseq_fastq_samples_excluded/
# Move back into directory to be tidy
mv colonocyte_RNAseq_fastq_samples_excluded colonocyte_RNAseq_fastq
```
Make sure to delete GMCC2S12_I3 on filezilla!
Repeat analysis
```{bash}
module load churp/0.2.2-slurm
# Make a directory in scratch space
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Work_samples_excluded
# Make metadata
$CHURP group_template bulk_rnaseq \
    -f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/colonocyte_RNAseq_fastq_samples_excluded \
    -o /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded/gmbc_ccyte_CHURP_meta.csv
```
```{bash}
$CHURP bulk_rnaseq \-e /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded/gmbc_ccyte_CHURP_meta.csv \-f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/colonocyte_RNAseq_fastq_samples_excluded \-x /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/hisat2/genome \-g /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/annotation/Homo_sapiens.GRCh38.100.gtf.gz \-o /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded \-d /scratch.global/sarif/gmbc_ccyte_CHURP/Work_samples_excluded \--rmdup --ppn 6 --mem 32000 -w 24 --no-submit
```
```{bash}
# Execute script
bash /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded/2022-04-24.sarif.bulk_rnaseq.pipeline.sh
```
Looking at the bulk rna seq report, here are the samples I want exlcuded from the merged files:
GMCC1S1_I1 (lots of rRNA contamination)
GMCC2S2_I3 (bad quality)
GMCC2S9_I3 (bad quality)
GMCC3S1_I3 (bad quality)
GMCC2S12_I3 (no reads)

First delete these samples from the samples_excluded directory (filezilla), then merge the remaining files.
```{bash}
#!/bin/sh        
#SBATCH --time=24:00:00
#SBATCH --ntasks=8
#SBATCH --mem=32g
#SBATCH --mail-type=ALL  
#SBATCH --mail-user=sarif@umn.edu

cd /panfs/roc/groups/7/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/colonocyte_RNAseq_fastq_samples_excluded

for i in $(ls -1 | sed 's/[_].*$//' | uniq)

    do echo "Merging R1"

cat "$i"_I*_R1_001.fastq.gz > "$i"_ME_R1_001.fastq.gz

       echo "Merging R2"

cat "$i"_I*_R2_001.fastq.gz > "$i"_ME_R2_001.fastq.gz

done;
```
Move files to new /merged subdirectory on filezilla, then repeat CHURP steps.
```{bash}
module load churp/0.2.2-slurm
# Make a directory in scratch space
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded_merged
mkdir -p /scratch.global/sarif/gmbc_ccyte_CHURP/Work_samples_excluded_merged
# Make metadata
$CHURP group_template bulk_rnaseq \
    -f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/colonocyte_RNAseq_fastq_samples_excluded/merged \
    -o /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded_merged/gmbc_ccyte_CHURP_meta.csv
```
```{bash}
$CHURP bulk_rnaseq \-e /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded_merged/gmbc_ccyte_CHURP_meta.csv \-f /home/blekhman/sarif/gmbc_ccyte/colonocyte_RNAseq_fastq/colonocyte_RNAseq_fastq_samples_excluded/merged \-x /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/hisat2/genome \-g /panfs/roc/risdb_new/ensembl/main/homo_sapiens/GRCh38/annotation/Homo_sapiens.GRCh38.100.gtf.gz \-o /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded_merged \-d /scratch.global/sarif/gmbc_ccyte_CHURP/Work_samples_excluded_merged \--rmdup --ppn 6 --mem 32000 -w 24 --no-submit
```
```{bash}
# Execute script
bash /scratch.global/sarif/gmbc_ccyte_CHURP/Out_samples_excluded_merged/2022-04-25.sarif.bulk_rnaseq.pipeline.sh
```

- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
SETUP

The relevant piece of information from the preprocessing steps is the counts matrix. I moved that from scratch to my MSI folder and also downloaded it to my home directory. 
```{r}
subread_cts <- read.table("data/RNAseq/merged_samplesexcluded/subread_counts_merged_samplesexcluded.txt", header=TRUE, sep="\t")
raw_read_cts <- as.matrix(subread_cts[,-c(1:6)])
# Gene ids to rownames
rownames(raw_read_cts) <- subread_cts[, 1] 
# Remove _ME string
colnames(raw_read_cts) <- gsub("_ME",  "", colnames(raw_read_cts), fixed = TRUE)
```
Load in technical metrics from sequencing, QC, mapping (merged samples only)
```{r}
# Extra technical metrics
meta_qc_additional <- read.csv("data/RNAseq/merged_samplesexcluded/RNASeq_Metrics_additional_merged_samplesexcluded.txt", sep="\t")
# Main files of technical metrics
meta_qc <- read.csv("data/RNAseq/merged_samplesexcluded/RNASeq_Metrics_merged_samplesexcluded.txt", sep="\t", header=F) %>%
  # Wide format and string removal
  reshape(idvar = "V1", timevar = "V2", direction = "wide") %>%
  rename_with(~str_remove(., 'V3.')) %>%
  # Add in extra technical metrics
  left_join(meta_qc_additional, by="V1") %>%
  # Column renaming
  rename_with(~str_replace_all(., " ", ".")) %>%
  mutate(SampleID = sub("_ME", "", V1)) %>%
  dplyr::rename(SampleID_CHURP = V1) %>%
  # Columns in alphabetical order
  dplyr::select(sort(names(.))) %>%
  relocate(SampleID, everything())
```
Load in main meta and add coculture condition 
```{r}
meta <- read.csv("data/metadata/GMCC_covariates_all_batches.csv") %>%
  mutate(SampleID = sub("_", "", SampleID)) %>%
  mutate(coculture = ifelse(grepl("control", SampleName), "control", "microbiome"))
```
Combine metadata with technical metrics (merged samples only)
```{r}
meta <- meta %>%
  left_join(meta_qc, by="SampleID")
write.csv(meta, "data/metadata/meta_RNAseq.csv")
```
Read count metrics
```{r}
# Minimum # reads
paste("min total:", min(meta_qc$Total.Reads))
paste("min trim:", min(meta_qc$Trimmed.R1.Count))
# Maximum # reads
paste("max total:", max(meta_qc$Total.Reads))
paste("max trim:", max(meta_qc$Trimmed.R1.Count))
# Average # reads
paste("avg total:", mean(meta_qc$Total.Reads))
paste("avg trim:", mean(meta_qc$Trimmed.R1.Count))
# Median # reads
paste("median total:", median(meta_qc$Total.Reads))
paste("median trim:", median(meta_qc$Trimmed.R1.Count))
# Average alignment rate
paste("avg align:", mean(meta_qc$Mapping.Rate))
```
After alignment, read counts ranged between 10,817,737 (15 million in sja) and 33,592,529 (56 million in sja) aligned reads per sample, with a mean of 17,142,585.72 (29 million in sja) and a median of 15,542,693.5 (27 million in sja) aligned reads per sample. Overall, the average alignment rate was ∼70% (89%) across samples

- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
TECHNICAL METRICS PLOTTING

Total reads (bar chart)
```{r}
ggplot(meta, aes(x=reorder(SampleID, Total.Reads), y=Total.Reads)) +
  geom_bar(stat="identity", fill="#b7094c") +
  #facet_wrap(~Batch, ncol=5, scales="free_x") +
  theme_linedraw() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
ggsave("results/figs/RNAseq/totalreads_RNA.png", width=5, height=2.5)

```
Mapping rate
```{r}
ggplot(meta, aes(x=reorder(SampleID, Mapping.Rate), y=Mapping.Rate)) +
  geom_bar(stat="identity", fill="#b7094c") +
  #facet_wrap(~Batch, ncol=5, scales="free_x") +
  ylim(values=c(0,1)) +
    theme_linedraw() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
ggsave("results/figs/RNAseq/mappingrate_RNA.png", width=5, height=2.5)
```

Read quality
```{r}
ggplot(meta, aes(x=SampleID, y=Average.Read.Quality, fill=Average.Read.Quality)) +
  geom_bar(stat="identity") +
  facet_wrap(~Batch, ncol=5, scales="free_x") + 
  binned_scale(aesthetics = "fill",
               scale_name = "stepsn", 
               palette = function(x) c("red", "orange", "pink", "yellow", "green", "forestgreen"),
               breaks = c(10, 20, 25, 30, 40, 45),
               limits = c(0, 45),
               show.limits = TRUE, 
               guide = "colorsteps"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

rRNA contamination
```{r}
ggplot(meta, aes(x=SampleID, y=rRNA_Pct, fill=rRNA_Pct)) +
  geom_bar(stat="identity") +
  facet_wrap(~Batch, ncol=5, scales="free_x") + 
  binned_scale(aesthetics = "fill",
               scale_name = "stepsn", 
               palette = function(x) c("grey40", "red"),
               breaks = c(0, 35),
               limits = c(0, 0.6),
               show.limits = TRUE, 
               guide = "colorsteps") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
4) % Fragments mapped to exons vs. introns vs. intergenic

5) Bar chart of uniquely vs. multiply vs discordantly vs unmapped reads

- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
FILTERING

First remove outlier samples
```{r}
outliers <- c("GMCC1S1", "GMCC4S10")
raw_read_cts <- raw_read_cts[, colnames(raw_read_cts) != outliers]
meta <- meta[meta$SampleID != outliers, ]
```
Next, use biomaRt to filter for only protein coding genes.
Select the human gene database:
```{r}
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", mirror=NULL)
```
Filter out non-protein coding genes and genes with no assigned novel genes (no symbol)
```{r}
# Run mart
raw_read_cts_anno <- biomaRt::getBM(
  attributes = c("ensembl_gene_id", "gene_biotype", "hgnc_symbol"), 
  filters = "ensembl_gene_id",
  values = rownames(raw_read_cts), 
  mart = ensembl)
```
```{r}
length(innate_immune_genes)
nrow(dplyr::filter(raw_read_cts_anno, ensembl_gene_id %in% innate_immune_genes))
```

```{r}
# Filter by protein coding
raw_read_cts_prtn <- raw_read_cts_anno %>%
  dplyr::filter(gene_biotype=="protein_coding") %>%
  dplyr::filter(!hgnc_symbol=="")
# Apply to counts matrix
raw_read_cts_prtn <- subset(raw_read_cts, rownames(raw_read_cts) %in% raw_read_cts_prtn$ensembl_gene_id)
# Check before and after
paste0("pre-filt: ", nrow(raw_read_cts))
paste0("post-filt: ", nrow(raw_read_cts_prtn))
```

Next, filter by abundance of zero-counts in groups
```{r}
# Create list of the SampleIDs corresponding to different groups
coculture_ctrl <- subset(meta$SampleID, meta$coculture == "control")
coculture_microbiome <- unlist(subset(meta$SampleID, meta$coculture == "microbiome"))
urbanism_urban <- unlist(subset(meta$SampleID, meta$urbanism == "urban"))
urbanism_rural <- unlist(subset(meta$SampleID, meta$urbanism == "rural"))

# Add columns to the counts matrix indicating the percentage of individuals with zero gene expression of each gene within a given group. 
raw_read_cts_grp <- as.data.frame(raw_read_cts_prtn)
raw_read_cts_grp <- raw_read_cts_grp %>%
  rownames_to_column('ensembl_gene_id') %>%
  mutate(perc0_coculture_ctrl = 
           rowSums(across(all_of(coculture_ctrl))==0)/length(coculture_ctrl)) %>%
  mutate(perc0_coculture_microbiome =
           rowSums(across(all_of(coculture_microbiome))==0)/length(coculture_microbiome)) %>%
  mutate(perc0_urbanism_urban = 
           rowSums(across(all_of(urbanism_urban))==0)/length(urbanism_urban)) %>%
  mutate(perc0_urbanism_rural = 
           rowSums(across(all_of(urbanism_rural))==0)/length(urbanism_rural))

# Keep genes which have 0 values in less than or equal to 50% control group samples OR 50% urban group samples OR 50% of rural group samples
raw_read_cts_filt <- raw_read_cts_grp %>%
  subset((perc0_coculture_ctrl <= 0.5 | perc0_coculture_microbiome <= 0.5)) %>%
  subset((perc0_urbanism_urban <= 0.5 | perc0_urbanism_rural <= 0.5)) %>%
  dplyr::select(-(starts_with("perc0"))) %>%
  `row.names<-`(., NULL) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()
write.csv(raw_read_cts_filt, "results/docs/raw_read_cts_filt.csv")

raw_read_cts_filt_alt <- raw_read_cts_grp %>%
  dplyr::filter(rowSums(dplyr::select(., -1)) >= 100) %>%
  `row.names<-`(.,NULL) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()
write.csv(raw_read_cts_filt, "results/docs/raw_read_cts_filt_alt.csv")


# Count difference
paste0("pre-filt: ", nrow(raw_read_cts_grp))
paste0("post-filt: ", nrow(raw_read_cts_filt))
paste0("post-filt: ", nrow(raw_read_cts_filt_alt))
```
```{r}
head(raw_read_cts_filt_alt)
```

Columns of all zeros create issues. Also, make sure continuous numeric columns are indeed numeric. Before, I had Total.Reads classified as a factor, meaning that every read was given its own column which obviously matched to only one sample, which produced a linear relationship which gave rise to the "not full rank" error.

Lastly, make sure all NAs are actually NA
```{r}
meta[meta == "N/A"] <- NA
```
