---
title: "R Notebook"
output: html_notebook
---

```{r}
library(ggplot2)
library(stringr)
library(pheatmap)
library(ensembldb)
library(biomaRt)
library(ggpubr)
library(tidyverse)
library(foreach)
library(DESeq2)
library(data.table)
library(dplyr)
```

--------------------------------------------------------------------------------
SETUP
--------------------------------------------------------------------------------

Load CLR-transformed abundances of microbes of interest
```{r}
# Read in CLR abundances of microbes
DA_microbe_abundances <- read.csv("results/docs/clr_filt_DA_microbe_abundances.csv", row.names = 1)
DA_microbe_abundances2 <- DA_microbe_abundances %>% rownames_to_column("SampleName")

# ALDEx2 data
aldex2_clr_all_sig <- read.csv("results/docs/DA_microbe_aldex2.csv", row.names = 1) %>%
  rownames_to_column("microbe")

# Attach to metadata
meta <- read.csv("data/metadata/meta_RNAseq.csv", row.names = 1) %>%
  left_join(DA_microbe_abundances2, by="SampleName") %>%
  # Batch as character
  mutate(Batch = as.character(Batch)) %>%
  # exclude controls
  dplyr::filter(!Luca_Lab_ID=="control") %>%
  # This sample was bad quality 
  dplyr::filter(!SampleID %in% c("GMCC1S1")) %>%
  # These samples were outliers amongst the RNAseq
  dplyr::filter(!SampleID %in% c("GMCC4S10", "GMCC3S7", "GMCC4C6", "GMCC5C5", "GMCC5C4", "GMCC5C3")) %>%
  # These samples were not used to treat the colonocytes
  dplyr::filter(!SampleName %in% c("1127XT", "1462QI", "3364PX", "3553XK", "3606IJ", "9423WC")) %>%
  # These samples did not have metagenomic sequencing data
  dplyr::filter(!SampleID %in% c("GMCC1S6", "GMCC2S3", "GMCC4S7"))
  #dplyr::filter(!SampleName=="4342XH")
meta <- meta[order(row.names(meta)), ]

DA_microbe_abundances2_melt <- DA_microbe_abundances2 %>%
  pivot_longer(
    cols = -c("SampleName"),
    names_to = "microbe",
    values_to = "microbe_abundance_CLR"
  ) %>%
  mutate(SampleName_microbe = paste(SampleName, microbe, sep = ";")) %>%
  left_join(aldex2_clr_all_sig, by="microbe")

# Covariates
covariates <- "Batch"
microbe_covariates_list <- paste(covariates, colnames(DA_microbe_abundances), sep = "+")
#microbe_covariates_list <- paste(covariates, microbe_list, sep = "+")
#microbe_covariates_list <- paste(microbe_covariates_list, "urbanism", sep = "+")
#microbe_covariates_list <- paste(microbe_covariates_list, microbe_list, sep = ":")

microbe_covariates_list <- paste("~", microbe_covariates_list) 
microbe_covariates_list <- lapply(microbe_covariates_list, function(x) as.formula(x))

names(microbe_covariates_list) <- colnames(DA_microbe_abundances)
```
Preview microbes we are testing
```{r}
ggplot(aldex2_clr_all_sig, aes(x=DA_urbanism)) +
  geom_bar(stat = "count") +
  xlab("Population in which microbe is differentially abundant") +
  ylab("Number of microbes") +
  ggtitle("Differentially abundant microbes (n=109)")
```
Effect sizes
```{r}
ggplot(aldex2_clr_all_sig, aes(x=DA_urbanism, y=abs(effect))) +
  geom_boxplot() +
  geom_jitter() +
  xlab("Population in which microbe is differentially abundant") +
  ylab("Effect size") +
  ggtitle("Differentially abundant microbes (n=109)")
```
Phylogenetic level
```{r}
aldex2_clr_all_sig$tax <- factor(aldex2_clr_all_sig$tax, levels=c("p", "c", "o", "f", "g", "s"))

ggplot(aldex2_clr_all_sig, aes(x=DA_urbanism, fill=tax)) +
  geom_bar(stat="count") +
  labs(fill = "Phylogenetic level") +
  xlab("Population in which microbe is differentially abundant") +
  ylab("Number of microbes") +
  ggtitle("Differentially abundant microbes (n=109)")
```

Load raw read counts
```{r}
raw_read_cts_filt <- read.csv("results/docs/raw_read_cts_filt.csv", row.names=1) 
raw_read_cts_filt <- raw_read_cts_filt[, colnames(raw_read_cts_filt) %in% meta$SampleID]
raw_read_cts_filt <- raw_read_cts_filt[, order(colnames(raw_read_cts_filt))]
```
Separate into urban-only and rural-only
```{r}
meta_urban <- meta %>% dplyr::filter(urbanism=="urban")
meta_urban <- meta_urban[order(row.names(meta_urban)), ]

meta_rural <- meta %>% dplyr::filter(urbanism=="rural")
meta_rural <- meta_rural[order(row.names(meta_rural)), ]

raw_read_cts_filt_urban <- raw_read_cts_filt[, colnames(raw_read_cts_filt) %in% meta_urban$SampleID]
raw_read_cts_filt_urban <- raw_read_cts_filt_urban[, order(colnames(raw_read_cts_filt_urban))]

raw_read_cts_filt_rural <- raw_read_cts_filt[, colnames(raw_read_cts_filt) %in% meta_rural$SampleID]
raw_read_cts_filt_rural <- raw_read_cts_filt_rural[, order(colnames(raw_read_cts_filt_rural))]
```


Option 1: Run locally

Function for running DESeq using each formula as a design
```{r}
DESeq_func_generaspecies <- function(x,y,z){
  DESeq(
    DESeqDataSetFromMatrix(
      countData = y,#raw_read_cts_filt,
      colData = z, #meta,
      design= x),
      quiet=T # silences progress messages; easier to see progress bar in the loop
  )}
```
edgeR for comparison
```{r}
library(edgeR)

# Assuming 'raw_read_cts_filt' is your count matrix and 'meta' is your metadata
# If not, replace them with your actual count matrix and metadata
edgeR_func <- function(counts_matrix, metadata_df, variable){
  
# Create a DGEList object
meta <- metadata_df
dge <- DGEList(counts = counts_matrix, group = meta$Batch)

# Filter low-abundance genes if needed
# You can adjust the threshold as per your data
keep <- rowSums(cpm(dge) > 1) >= 3
dge <- dge[keep, ]

# Define design matrix
design <- model.matrix(variable, data = meta)

# Estimate dispersion
dge <- estimateDisp(dge, design)

# Fit the generalized linear model
fit <- glmFit(dge, design)

# Perform likelihood ratio test
lrt <- glmLRT(fit, coef = 2)

# Extract results
results <- topTags(lrt, n = nrow(dge))$table

# Adjust p-values if needed
# results$PValue <- p.adjust(results$PValue, method = "BH")

# Print the results
return(results)
}

```


First set up a parallel backend so the models can be run in parallel. 
Source: https://www.blasbenito.com/post/02_parallelizing_loops_with_r/
```{r}
# Leave one core free for other tasks
n.cores <- parallel::detectCores() - 1
#create the cluster
my.cluster <- parallel::makeCluster(
  n.cores, 
  type = "PSOCK",
  outfile=""
  )
#register it to be used by %dopar%
library(doParallel)
doParallel::registerDoParallel(cl = my.cluster)
```
Iterate through every design when performing DESeq: output as individual DESeq objects in a list. Skip models that error. Although this loop is run in parallel, it still takes awhile. Other loops we run shouldn't need to be run in parallel.
```{r}
x <- microbe_covariates_list
dds_list <- foreach(
  i=seq_along(x),
  .packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- DESeq_func_generaspecies(x[[i]], raw_read_cts_filt, meta)},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    } 
```
edgeR
```{r}
x <- microbe_covariates_list
edgeR_list <- foreach(
  i=seq_along(x),
  .packages='edgeR',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- edgeR_func(raw_read_cts_filt, meta, x[[i]])},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    } 
```
Urban only
```{r}
x <- microbe_covariates_list
dds_list_urban <- foreach(
  i=seq_along(x),
  .packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- DESeq_func_generaspecies(x[[i]], raw_read_cts_filt_urban, meta_urban)},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    } 
```
edgeR
```{r}
x <- microbe_covariates_list
edgeR_list_urban <- foreach(
  i=seq_along(x),
  .packages='edgeR',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- edgeR_func(raw_read_cts_filt_urban, meta_urban, x[[i]])},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    } 
```


Rural only
```{r}
x <- microbe_covariates_list
dds_list_rural <- foreach(
  i=seq_along(x),
  .packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- DESeq_func_generaspecies(x[[i]], raw_read_cts_filt_rural, meta_rural)},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    } 
```
edgeR
```{r}
x <- microbe_covariates_list
edgeR_list_rural <- foreach(
  i=seq_along(x),
  .packages='edgeR',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- edgeR_func(raw_read_cts_filt_rural, meta_rural, x[[i]])},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    } 
```
```{r}
# Save RDS object
saveRDS(dds_list, "results/rds/dds_list.rds")
saveRDS(dds_list_urban, "results/rds/dds_list_urban.rds")
saveRDS(dds_list_rural, "results/rds/dds_list_rural.rds")
```



Option 2: Run on HPC
First save everything to file
```{r}
#saveRDS(DESeq_func_generaspecies, "data/for_hpc/DESeq_func_generaspecies.rds")
#saveRDS(microbe_covariates_list, "data/for_hpc/microbe_covariates_list_CSS_BATCH.rds")
#saveRDS(raw_read_cts_filt, "data/for_hpc/raw_read_cts_filt.rds")
#saveRDS(meta, "data/for_hpc/meta_CSS.rds")
```
rsync every .rds file in the for_hpc folder (can run directly from here)
```{bash}
scp -r data/for_hpc/*.rds sjarif@midway3.rcc.uchicago.edu:gmbc_colonocytes/05_RNAseq-metaG-microbe/data/
```
Now create R script and save
```{R}
library(DESeq2)
library(tidyverse)
library(foreach)
library(doParallel)

# make sure this is same as number of cpus requested
registerDoParallel(cores=8)

DESeq_func_generaspecies <- readRDS("../data/DESeq_func_generaspecies.rds")
microbe_covariates_list <- readRDS("../data/microbe_covariates_list.rds")
raw_read_cts_filt <- readRDS("../data/raw_read_cts_filt.rds")
meta <- readRDS("../data/meta.rds")


x <- microbe_covariates_list
dds_list <- foreach(
  i=seq_along(x),
  .packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %dopar% {
    tryCatch({
      y <- DESeq_func_generaspecies(x[[i]], raw_read_cts_filt, meta)},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  } 
  
saveRDS(dds_list, "dds_list.rds")
```
Create bash file to run R script
```{bash}
#!/bin/bash      
#SBATCH --time=24:00:00
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8
#SBATCH --mem=60g
#SBATCH -p blekhman
#SBATCH --account pi-blekhman
#SBATCH --mail-type=ALL  
#SBATCH --mail-user=sjarif@uchicago.edu

cd /home/sjarif/gmbc_colonocytes/05_RNAseq-metaG-microbe/results

# Load the necessary modules if needed
module load R

# Run the R script using Rscript
Rscript ../code/dds.R
```
Check the resources that were used. According to this it took 12 minutes and used at max 16G RAM
```{bash}
# JUST AN EXAMPLE
sacct -j 9894616 --format=JobID,JobName,MaxRSS,Elapsed,UserCPU
```
Download and proceed as normal (better to run this in terminal from local drive so you can see progress)
```{bash}
scp -r sjarif@midway3.rcc.uchicago.edu:gmbc_colonocytes/05_RNAseq-metaG-microbe/results/dds_list_CLR_BATCH_URBANISM.rds /Users/sabri/Documents/GitHub/gmbc-colonocytes/results/rds

scp -r sjarif@midway3.rcc.uchicago.edu:gmbc_colonocytes/05_RNAseq-metaG-microbe/results/dds_list_CSS_BATCH_URBANISM.rds /Users/sabri/Documents/GitHub/gmbc-colonocytes/results/rds

scp -r sjarif@midway3.rcc.uchicago.edu:gmbc_colonocytes/05_RNAseq-metaG-microbe/results/dds_list_CLR_BATCH_URBANISM_INTERACTION.rds /Users/sabri/Documents/GitHub/gmbc-colonocytes/results/rds

scp -r sjarif@midway3.rcc.uchicago.edu:gmbc_colonocytes/05_RNAseq-metaG-microbe/results/dds_list_CSS_BATCH_URBANISM_INTERACTION.rds /Users/sabri/Documents/GitHub/gmbc-colonocytes/results/rds
```











Back to local
```{r}
dds_list <- readRDS("results/rds/dds_list.rds")
dds_list_urban <- readRDS("results/rds/dds_list_urban.rds")
dds_list_rural <- readRDS("results/rds/dds_list_rural.rds")
```
Examine cooks
```{r}
maxCooks <- apply(assays(dds_list_rural[[1]])[["cooks"]], 1, max)
head(maxCooks)
```

```{r}
ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl", host="jul2023.archive.ensembl.org")

# Convert DESeq results into table
res2table <- function(ddsobject){
  results(ddsobject, cooksCutoff=T)  %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") }

# Define annotations
att_interest <- c("ensembl_gene_id", "description", "hgnc_symbol") #"ensembl_transcript_id", 

# Annotate 
annotate_ensembl <- function(resdf){biomaRt::getBM(
  attributes = att_interest, 
  filters = "ensembl_gene_id",
  values = resdf$ensembl_gene_id, 
  mart = ensembl) %>%
  left_join(resdf, by="ensembl_gene_id")
}
```
Loop through list of DESeq objects and generate result tables
```{r}
dds_list_filt <- Filter(Negate(is.null), dds_list)

x <- dds_list_filt
res_list <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <- res2table(x[[i]])},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }
```
Urban only
```{r}
dds_list_filt_urban <- Filter(Negate(is.null), dds_list_urban)

x <- dds_list_filt_urban
res_list_urban <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <- res2table(x[[i]])},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }
```
Rural only
```{r}
dds_list_filt_rural <- Filter(Negate(is.null), dds_list_rural)

x <- dds_list_filt_rural
res_list_rural <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <- res2table(x[[i]])},
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }
```

Filter for genes with an FDR <= 0.1
```{r}
x <- res_list
sig_list <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], padj < 0.1 ) 
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

sig_list_filt <- Filter(NROW, sig_list)
```
edgeR
```{r}
x <- edgeR_list
edgeR_sig_list <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], FDR < 0.1 ) 
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

edgeR_sig_list_filt <- Filter(NROW, edgeR_sig_list)
```
Urban only
```{r}
x <- res_list_urban
sig_list_urban <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], padj < 0.1 ) 
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

sig_list_filt_urban <- Filter(NROW, sig_list_urban)
```
edgeR
```{r}
x <- edgeR_list_urban
edgeR_sig_list_urban <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], FDR < 0.1 ) 
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

edgeR_sig_list_filt_urban <- Filter(NROW, edgeR_sig_list_urban)
```
Rural only
```{r}
x <- res_list_rural
sig_list_rural <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], padj < 0.1 ) 
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

sig_list_filt_rural <- Filter(NROW, sig_list_rural)
```
edgeR
```{r}
x <- edgeR_list_rural
edgeR_sig_list_rural <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], FDR < 0.1 ) 
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

edgeR_sig_list_filt_rural <- Filter(NROW, edgeR_sig_list_rural)
```

Summary
```{r}
cat("all colonocytes - number of microbes", "\n")
cat("deseq:", length(sig_list_filt), "  edgeR:", length(edgeR_sig_list_filt), "\n")
cat("urban colonocytes - number of microbes", "\n")
cat("deseq:", length(sig_list_filt_urban), "   edgeR:", length(edgeR_sig_list_filt_urban), "\n")
cat("rural colonocytes - number of microbes", "\n")
cat("deseq:", length(sig_list_filt_rural), "   edgeR:", length(edgeR_sig_list_filt_rural))
```
```{r}
cat("all colonocytes - shared microbes", "\n")
cat("unique to deseq:", length(setdiff(names(sig_list_filt), names(edgeR_sig_list_filt))),
    "  unique to edgeR:", length(setdiff(names(edgeR_sig_list_filt), names(sig_list_filt))),
    "  shared:", length(intersect(names(sig_list_filt), names(edgeR_sig_list_filt))), "\n")

cat("urban colonocytes - shared microbes", "\n")
cat("unique to deseq:", length(setdiff(names(sig_list_filt_urban), names(edgeR_sig_list_filt_urban))),
    "  unique to edgeR:", length(setdiff(names(edgeR_sig_list_filt_urban), names(sig_list_filt_urban))),
    "  shared:", length(intersect(names(sig_list_filt_urban), names(edgeR_sig_list_filt_urban))), "\n")

cat("rural colonocytes - shared microbes", "\n")
cat("unique to deseq:", length(setdiff(names(sig_list_filt_rural), names(edgeR_sig_list_filt_rural))),
    "  unique to edgeR:", length(setdiff(names(edgeR_sig_list_filt_rural), names(sig_list_filt_rural))),
    "  shared:", length(intersect(names(sig_list_filt_rural), names(edgeR_sig_list_filt_rural))), "\n")
```


Append model name to each dataframe
```{r}
x <- sig_list_filt
sig_list_filt_names <- foreach(
  i=seq_along(x),
  n=names(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::mutate(x[[i]], microbe = n) %>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

```
Urban only
```{r}
x <- sig_list_filt_urban
sig_list_filt_names_urban <- foreach(
  i=seq_along(x),
  n=names(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::mutate(x[[i]], microbe = n) %>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

```
Rural only
```{r}
x <- sig_list_filt_rural
sig_list_filt_names_rural <- foreach(
  i=seq_along(x),
  n=names(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::mutate(x[[i]], microbe = n) %>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

```

Flatten this list of dfs
```{r}
sig_list_df <- bind_rows(sig_list_filt_names) %>%
  dplyr::select(ensembl_gene_id, microbe, everything()) %>%
  mutate(l2fc_dir = case_when(log2FoldChange > 0 ~ "U",
                               log2FoldChange < 0 ~ "D")) %>%
  left_join(aldex2_clr_all_sig[c("microbe", "DA_urbanism")], by="microbe")
```

Urban only
```{r}
sig_list_df_urban <- bind_rows(sig_list_filt_names_urban) %>%
  dplyr::select(ensembl_gene_id, microbe, everything()) %>%
  mutate(l2fc_dir = case_when(log2FoldChange > 0 ~ "U",
                               log2FoldChange < 0 ~ "D")) %>%
  left_join(aldex2_clr_all_sig[c("microbe", "DA_urbanism")], by="microbe")
```
Rural only
```{r}
sig_list_df_rural <- bind_rows(sig_list_filt_names_rural) %>%
  dplyr::select(ensembl_gene_id, microbe, everything()) %>%
  mutate(l2fc_dir = case_when(log2FoldChange > 0 ~ "U",
                               log2FoldChange < 0 ~ "D")) %>%
  left_join(aldex2_clr_all_sig[c("microbe", "DA_urbanism")], by="microbe")
```


Summary df: rows are microbes
```{r}
# New df; fill will empty rows that correspond to number of models
sig_list_summary <- data.frame(n_genes=numeric(0)) 
sig_list_summary[nrow(sig_list_summary)+length(sig_list_filt),] <- NA
# Assign rownames based on sigres_list
rownames(sig_list_summary) <- names(sig_list_filt)
sig_list_summary$microbe <- row.names(sig_list_summary)
# Loop through each sig list
sig_list_summary$n_genes <- as.numeric(lapply(sig_list_filt, function(x) length(unique(x[["ensembl_gene_id"]]))))
# list unique gene IDs
sig_list_summary$genes <- lapply(sig_list_filt, function(x) str_c(unique(x[["ensembl_gene_id"]]), collapse = ";"))

# Step 1: Merge generaspecies_sig_list_summary with sig_list_df to obtain log2FoldChange values
merged_df <- merge(sig_list_summary, sig_list_df, by = "microbe")

# Step 2: Group the merged dataframe by "model" and "ensembl_gene_id" and count the number of genes that are upregulated and downregulated
#detach(package:plyr)
counts_df <- merged_df %>%
  group_by(microbe) %>%
  summarise(n_U = sum(l2fc_dir == "U"),
            n_D = sum(l2fc_dir == "D")) 

# Step 4: Merge the summary dataframe with generaspecies_sig_list_summary to add the n_U and n_D columns
sig_list_summary <- merge(sig_list_summary, counts_df, by = "microbe") %>%
  left_join(aldex2_clr_all_sig[c("microbe", "DA_urbanism")], by="microbe")

```
Urban only
```{r}
# New df; fill will empty rows that correspond to number of models
sig_list_summary_urban <- data.frame(n_genes=numeric(0)) 
sig_list_summary_urban[nrow(sig_list_summary_urban)+length(sig_list_filt_urban),] <- NA
# Assign rownames based on sigres_list
rownames(sig_list_summary_urban) <- names(sig_list_filt_urban)
sig_list_summary_urban$microbe <- row.names(sig_list_summary_urban)
# Loop through each sig list
sig_list_summary_urban$n_genes <- as.numeric(lapply(sig_list_filt_urban, function(x) length(unique(x[["ensembl_gene_id"]]))))
# list unique gene IDs
sig_list_summary_urban$genes <- lapply(sig_list_filt_urban, function(x) str_c(unique(x[["ensembl_gene_id"]]), collapse = ";"))

# Step 1: Merge generaspecies_sig_list_summary with sig_list_df to obtain log2FoldChange values
merged_df_urban <- merge(sig_list_summary_urban, sig_list_df_urban, by = "microbe")

# Step 2: Group the merged dataframe by "model" and "ensembl_gene_id" and count the number of genes that are upregulated and downregulated
#detach(package:plyr)
counts_df_urban <- merged_df_urban %>%
  group_by(microbe) %>%
  summarise(n_U = sum(l2fc_dir == "U"),
            n_D = sum(l2fc_dir == "D")) 

# Step 4: Merge the summary dataframe with generaspecies_sig_list_summary to add the n_U and n_D columns
sig_list_summary_urban <- merge(sig_list_summary_urban, counts_df_urban, by = "microbe") %>%
  left_join(aldex2_clr_all_sig[c("microbe", "DA_urbanism")], by="microbe")

```
Rural only
```{r}
# New df; fill will empty rows that correspond to number of models
sig_list_summary_rural <- data.frame(n_genes=numeric(0)) 
sig_list_summary_rural[nrow(sig_list_summary_rural)+length(sig_list_filt_rural),] <- NA
# Assign rownames based on sigres_list
rownames(sig_list_summary_rural) <- names(sig_list_filt_rural)
sig_list_summary_rural$microbe <- row.names(sig_list_summary_rural)
# Loop through each sig list
sig_list_summary_rural$n_genes <- as.numeric(lapply(sig_list_filt_rural, function(x) length(unique(x[["ensembl_gene_id"]]))))
# list unique gene IDs
sig_list_summary_rural$genes <- lapply(sig_list_filt_rural, function(x) str_c(unique(x[["ensembl_gene_id"]]), collapse = ";"))

# Step 1: Merge generaspecies_sig_list_summary with sig_list_df to obtain log2FoldChange values
merged_df_rural <- merge(sig_list_summary_rural, sig_list_df_rural, by = "microbe")

# Step 2: Group the merged dataframe by "model" and "ensembl_gene_id" and count the number of genes that are upregulated and downregulated
#detach(package:plyr)
counts_df_rural <- merged_df_rural %>%
  group_by(microbe) %>%
  summarise(n_U = sum(l2fc_dir == "U"),
            n_D = sum(l2fc_dir == "D")) 

# Step 4: Merge the summary dataframe with generaspecies_sig_list_summary to add the n_U and n_D columns
sig_list_summary_rural <- merge(sig_list_summary_rural, counts_df_rural, by = "microbe") %>%
  left_join(aldex2_clr_all_sig[c("microbe", "DA_urbanism")], by="microbe")

```


Summary df: rows are genes
```{r}
gene_list_summary <- sig_list_df %>%
  group_by(hgnc_symbol) %>%
  summarise(
    n_microbes = n(),
            urban_U = sum(DA_urbanism == "urban" & l2fc_dir=="U"),
            urban_D = sum(DA_urbanism == "urban" & l2fc_dir=="D"),
            rural_U = sum(DA_urbanism == "rural" & l2fc_dir=="U"),
            rural_D= sum(DA_urbanism == "rural" & l2fc_dir=="D"))

gene_list_summary_plot <- gene_list_summary %>%
  mutate(hgnc_symbol = factor(hgnc_symbol, levels = unique(hgnc_symbol))) %>%
  #dplyr::filter(n_microbes >= 10) %>%
  pivot_longer(cols = c(urban_U, urban_D, rural_U, rural_D),
               names_to = c("DA_urbanism", "l2fc_dir"),
               names_pattern = "(.*?)_(.*)",
               values_to = "count")
```
Urban only
```{r}
gene_list_summary_urban <- sig_list_df_urban %>%
  #dplyr::filter(!microbe %in% c("Alistipes_megaguti", "Alistipes_communis", "Bacteroides_faecichinchillae", "Erysipelatoclostridium", "Unknown_genus_70")) %>%
  group_by(hgnc_symbol) %>%
  summarise(
    n_microbes = n(),
            urban_U = sum(DA_urbanism == "urban" & l2fc_dir=="U"),
            urban_D = sum(DA_urbanism == "urban" & l2fc_dir=="D"),
            rural_U = sum(DA_urbanism == "rural" & l2fc_dir=="U"),
            rural_D= sum(DA_urbanism == "rural" & l2fc_dir=="D"))

gene_list_summary_plot_urban <- gene_list_summary_urban %>%
  mutate(hgnc_symbol = factor(hgnc_symbol, levels = unique(hgnc_symbol))) %>%
  #dplyr::filter(n_microbes >= 10) %>%
  pivot_longer(cols = c(urban_U, urban_D, rural_U, rural_D),
               names_to = c("DA_urbanism", "l2fc_dir"),
               names_pattern = "(.*?)_(.*)",
               values_to = "count")
```
Rural only
```{r}
gene_list_summary_rural <- sig_list_df_rural %>%
  #dplyr::filter(!microbe %in% c("Unknown_genus_54", "Holdemanella_sp_2", "Parasutterella_excrementihominis", "Parasutterella", "Erysipelatoclostridium", "Moraxellaceae", "Acinetobacter", "Acinetobacter_baumannii", "Holdemanella_biformis", "Clostridium_innocuum")) %>%
  group_by(hgnc_symbol) %>%
  summarise(
    n_microbes = n(),
            urban_U = sum(DA_urbanism == "urban" & l2fc_dir=="U"),
            urban_D = sum(DA_urbanism == "urban" & l2fc_dir=="D"),
            rural_U = sum(DA_urbanism == "rural" & l2fc_dir=="U"),
            rural_D= sum(DA_urbanism == "rural" & l2fc_dir=="D"))

gene_list_summary_plot_rural <- gene_list_summary_rural %>%
  mutate(hgnc_symbol = factor(hgnc_symbol, levels = unique(hgnc_symbol))) %>%
  #dplyr::filter(n_microbes >= 10) %>%
  pivot_longer(cols = c(urban_U, urban_D, rural_U, rural_D),
               names_to = c("DA_urbanism", "l2fc_dir"),
               names_pattern = "(.*?)_(.*)",
               values_to = "count")
```

```{r}
gene_list_summary_plot <- gene_list_summary_plot %>%
  dplyr::arrange(n_microbes, hgnc_symbol) %>%
  mutate(count = as.numeric(count))

gene_list_summary_plot$hgnc_symbol <- factor(gene_list_summary_plot$hgnc_symbol, levels = unique(gene_list_summary_plot$hgnc_symbol))

ggplot(dplyr::filter(gene_list_summary_plot, gene_list_summary_plot$l2fc_dir=="U"), aes(x = DA_urbanism, y = hgnc_symbol, color=DA_urbanism)) + 
  geom_point(aes(size = ifelse(count == 0, NA, count))) +
  scale_size_continuous(limits = c(0, 50)) +
  labs(color = "Population in which microbe is differentially abundant") +
  labs(size = "Number of microbes") + # Optional: Label for the size legend
  scale_color_manual(values=c("#6FBD8B", "#7B2BC1")) +
  facet_grid(~l2fc_dir) +
  theme_minimal() +
  scale_x_discrete(labels = c("Rural", "Urban")) +
  theme(strip.background = element_blank(),
        #strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", hjust=0.5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.2, "cm"))
#ggsave("results/figs/RNAseq-metaG-microbe/gene_diffabun_heatmap_U.png", width=3, height=6)

ggplot(dplyr::filter(gene_list_summary_plot, gene_list_summary_plot$l2fc_dir=="D"), aes(x = DA_urbanism, y = hgnc_symbol, color=DA_urbanism)) + 
  geom_point(aes(size = ifelse(count == 0, NA, count))) +
  scale_size_continuous(limits = c(0, 50)) +
  labs(color = "Population in which microbe is differentially abundant") +
  labs(size = "Number of microbes") + # Optional: Label for the size legend
  scale_color_manual(values=c("#6FBD8B", "#7B2BC1")) +
  facet_grid(~l2fc_dir) +
  theme_minimal() +
  scale_x_discrete(labels = c("Rural", "Urban")) +
  theme(strip.background = element_blank(),
        #strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", hjust=0.5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.2, "cm"))
#ggsave("results/figs/RNAseq-metaG-microbe/gene_diffabun_heatmap_D.png", width=3, height=6)
```
```{r}
gene_list_summary_plot_urban <- gene_list_summary_plot_urban %>%
  dplyr::arrange(n_microbes, hgnc_symbol) %>%
  mutate(count = as.numeric(count))

gene_list_summary_plot_urban$hgnc_symbol <- factor(gene_list_summary_plot_urban$hgnc_symbol, levels = unique(gene_list_summary_plot_urban$hgnc_symbol))

ggplot(dplyr::filter(gene_list_summary_plot_urban, gene_list_summary_plot_urban$l2fc_dir=="U"), aes(x = DA_urbanism, y = hgnc_symbol, color=DA_urbanism)) + 
  geom_point(aes(size = ifelse(count == 0, NA, count))) +
  scale_size_continuous(limits = c(0, 5)) +
  labs(color = "Population in which microbe is differentially abundant") +
  labs(size = "Number of microbes") + # Optional: Label for the size legend
  scale_color_manual(values=c("#6FBD8B", "#7B2BC1")) +
  facet_grid(~l2fc_dir) +
  theme_minimal() +
  scale_x_discrete(labels = c("Rural", "Urban")) +
  theme(strip.background = element_blank(),
        #strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", hjust=0.5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.2, "cm"))
#ggsave("results/figs/RNAseq-metaG-microbe/gene_diffabun_heatmap_U.png", width=3, height=6)

ggplot(dplyr::filter(gene_list_summary_plot_urban, gene_list_summary_plot_urban$l2fc_dir=="D"), aes(x = DA_urbanism, y = hgnc_symbol, color=DA_urbanism)) + 
  geom_point(aes(size = ifelse(count == 0, NA, count))) +
  scale_size_continuous(limits = c(0, 5)) +
  labs(color = "Population in which microbe is differentially abundant") +
  labs(size = "Number of microbes") + # Optional: Label for the size legend
  scale_color_manual(values=c("#6FBD8B", "#7B2BC1")) +
  facet_grid(~l2fc_dir) +
  theme_minimal() +
  scale_x_discrete(labels = c("Rural", "Urban")) +
  theme(strip.background = element_blank(),
        #strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", hjust=0.5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.2, "cm"))
#ggsave("results/figs/RNAseq-metaG-microbe/gene_diffabun_heatmap_D.png", width=3, height=6)
```
```{r}
gene_list_summary_plot_rural <- gene_list_summary_plot_rural %>%
  dplyr::arrange(n_microbes, hgnc_symbol) %>%
  mutate(count = as.numeric(count))

gene_list_summary_plot_rural$hgnc_symbol <- factor(gene_list_summary_plot_rural$hgnc_symbol, levels = unique(gene_list_summary_plot_rural$hgnc_symbol))

ggplot(dplyr::filter(gene_list_summary_plot_rural, gene_list_summary_plot_rural$l2fc_dir=="U"), aes(x = DA_urbanism, y = hgnc_symbol, color=DA_urbanism)) + 
  geom_point(aes(size = ifelse(count == 0, NA, count))) +
  scale_size_continuous(limits = c(0, 5)) +
  labs(color = "Population in which microbe is differentially abundant") +
  labs(size = "Number of microbes") + # Optional: Label for the size legend
  scale_color_manual(values=c("#6FBD8B", "#7B2BC1")) +
  facet_grid(~l2fc_dir) +
  theme_minimal() +
  scale_x_discrete(labels = c("Rural", "Urban")) +
  theme(strip.background = element_blank(),
        #strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", hjust=0.5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.2, "cm"))
#ggsave("results/figs/RNAseq-metaG-microbe/gene_diffabun_heatmap_U.png", width=3, height=6)

ggplot(dplyr::filter(gene_list_summary_plot_rural, gene_list_summary_plot_rural$l2fc_dir=="D"), aes(x = DA_urbanism, y = hgnc_symbol, color=DA_urbanism)) + 
  geom_point(aes(size = ifelse(count == 0, NA, count))) +
  scale_size_continuous(limits = c(0, 5)) +
  labs(color = "Population in which microbe is differentially abundant") +
  labs(size = "Number of microbes") + # Optional: Label for the size legend
  scale_color_manual(values=c("#6FBD8B", "#7B2BC1")) +
  facet_grid(~l2fc_dir) +
  theme_minimal() +
  scale_x_discrete(labels = c("Rural", "Urban")) +
  theme(strip.background = element_blank(),
        #strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", hjust=0.5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.2, "cm"))
#ggsave("results/figs/RNAseq-metaG-microbe/gene_diffabun_heatmap_D.png", width=3, height=6)
```
How to derive residuals: https://support.bioconductor.org/p/60567/
```{r}
# Extract the residuals (batch and total.reads corrected counts)
x <- dds_list_filt
residuals <- foreach(
  i=seq_along(x),
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      fitted.common.scale = t(t(assays(x[[i]])[["mu"]])/sizeFactors(x[[i]]))
      y <-  counts(x[[i]], normalized=T) - fitted.common.scale
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

# Filter for microbes of interest
residuals_filt1 <- residuals[names(sig_list_filt_names)]

# Bind the matching sig_list_filt
combine_function <- function(item1, item2) {
  new_item <- list(residuals = item1, sig_results = item2)
  return(new_item)
}
residuals_paired <- Map(combine_function, residuals_filt1, sig_list_filt_names)

# function to filter by significant genes
filter_residuals <- function(residuals, sig_results) {
  filtered_residuals <- residuals[sig_results$ensembl_gene_id, , drop = FALSE]
  return(filtered_residuals)
}

# Use lapply to filter the "residuals" sub-element and retain "sig_results"
residuals_filt2 <- lapply(residuals_paired, function(paired_element) {
  filtered_residuals <- filter_residuals(paired_element$residuals, paired_element$sig_results) %>% 
    as.data.frame() %>% 
    rownames_to_column("ensembl_gene_id") %>%
    left_join(paired_element$sig_results[c("ensembl_gene_id", "microbe", "hgnc_symbol")], by="ensembl_gene_id") %>%
    mutate(microbe_gene = paste(microbe, hgnc_symbol, sep = ";")) 
  paired_element$residuals <- filtered_residuals  # Update the "residuals" sub-element
  return(paired_element$residuals)  # Return the modified element with "sig_results" intact
})

residuals_rbind <- as.data.frame(do.call(rbind, residuals_filt2)) %>%
  dplyr::select(microbe_gene, everything())

residuals_melt <- residuals_rbind  %>%
  pivot_longer(
    cols = -c("microbe_gene", "ensembl_gene_id", "microbe", "hgnc_symbol"),
    names_to = "SampleID",
    values_to = "residual_gene_count"
  ) %>%
  left_join(meta[c("SampleName", "SampleID")], by = "SampleID") %>%
  mutate(SampleName_microbe = paste(SampleName, microbe, sep = ";")) %>%
  left_join(DA_microbe_abundances2_melt[c("SampleName_microbe", "microbe_abundance_CLR")], by="SampleName_microbe") %>%
  dplyr::select(c("microbe_gene", "microbe_abundance_CLR", "residual_gene_count"))
```
Urban only
```{r}
# Extract the residuals (batch and total.reads corrected counts)
x <- dds_list_filt_urban
residuals_urban <- foreach(
  i=seq_along(x),
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      fitted.common.scale = t(t(assays(x[[i]])[["mu"]])/sizeFactors(x[[i]]))
      y <-  counts(x[[i]], normalized=T) - fitted.common.scale
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

# Filter for microbes of interest
residuals_filt1_urban <- residuals_urban[names(sig_list_filt_names_urban)]

# Bind the matching sig_list_filt
combine_function <- function(item1, item2) {
  new_item <- list(residuals = item1, sig_results = item2)
  return(new_item)
}
residuals_paired_urban <- Map(combine_function, residuals_filt1_urban, sig_list_filt_names_urban)

# function to filter by significant genes
filter_residuals <- function(residuals, sig_results) {
  filtered_residuals <- residuals[sig_results$ensembl_gene_id, , drop = FALSE]
  return(filtered_residuals)
}

# Use lapply to filter the "residuals" sub-element and retain "sig_results"
residuals_filt2_urban <- lapply(residuals_paired_urban, function(paired_element) {
  filtered_residuals <- filter_residuals(paired_element$residuals, paired_element$sig_results) %>% 
    as.data.frame() %>% 
    rownames_to_column("ensembl_gene_id") %>%
    left_join(paired_element$sig_results[c("ensembl_gene_id", "microbe", "hgnc_symbol")], by="ensembl_gene_id") %>%
    mutate(microbe_gene = paste(microbe, hgnc_symbol, sep = ";")) 
  paired_element$residuals <- filtered_residuals  # Update the "residuals" sub-element
  return(paired_element$residuals)  # Return the modified element with "sig_results" intact
})

residuals_rbind_urban <- as.data.frame(do.call(rbind, residuals_filt2_urban)) %>%
  dplyr::select(microbe_gene, everything())

residuals_melt_urban <- residuals_rbind_urban  %>%
  pivot_longer(
    cols = -c("microbe_gene", "ensembl_gene_id", "microbe", "hgnc_symbol"),
    names_to = "SampleID",
    values_to = "residual_gene_count"
  ) %>%
  dplyr::filter(!microbe %in% c("Erysipelotrichaceae", "Erysipelotrichales", "Erysipelotrichia", "Unknown_genus_82")) %>%
  left_join(meta_urban[c("SampleName", "SampleID")], by = "SampleID") %>%
  mutate(SampleName_microbe = paste(SampleName, microbe, sep = ";")) %>%
  left_join(DA_microbe_abundances2_melt[c("SampleName_microbe", "microbe_abundance_CLR")], by="SampleName_microbe") %>%
  dplyr::select(c("microbe_gene", "microbe_abundance_CLR", "residual_gene_count"))
```
Rural only
```{r}
# Extract the residuals (batch and total.reads corrected counts)
x <- dds_list_filt_rural
residuals_rural <- foreach(
  i=seq_along(x),
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      fitted.common.scale = t(t(assays(x[[i]])[["mu"]])/sizeFactors(x[[i]]))
      y <-  counts(x[[i]], normalized=T) - fitted.common.scale
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

# Filter for microbes of interest
residuals_filt1_rural <- residuals_rural[names(sig_list_filt_names_rural)]

# Bind the matching sig_list_filt
combine_function <- function(item1, item2) {
  new_item <- list(residuals = item1, sig_results = item2)
  return(new_item)
}
residuals_paired_rural <- Map(combine_function, residuals_filt1_rural, sig_list_filt_names_rural)

# function to filter by significant genes
filter_residuals <- function(residuals, sig_results) {
  filtered_residuals <- residuals[sig_results$ensembl_gene_id, , drop = FALSE]
  return(filtered_residuals)
}

# Use lapply to filter the "residuals" sub-element and retain "sig_results"
residuals_filt2_rural <- lapply(residuals_paired_rural, function(paired_element) {
  filtered_residuals <- filter_residuals(paired_element$residuals, paired_element$sig_results) %>% 
    as.data.frame() %>% 
    rownames_to_column("ensembl_gene_id") %>%
    left_join(paired_element$sig_results[c("ensembl_gene_id", "microbe", "hgnc_symbol")], by="ensembl_gene_id") %>%
    mutate(microbe_gene = paste(microbe, hgnc_symbol, sep = ";")) 
  paired_element$residuals <- filtered_residuals  # Update the "residuals" sub-element
  return(paired_element$residuals)  # Return the modified element with "sig_results" intact
})

residuals_rbind_rural <- as.data.frame(do.call(rbind, residuals_filt2_rural)) %>%
  dplyr::select(microbe_gene, everything())

residuals_melt_rural <- residuals_rbind_rural  %>%
  pivot_longer(
    cols = -c("microbe_gene", "ensembl_gene_id", "microbe", "hgnc_symbol"),
    names_to = "SampleID",
    values_to = "residual_gene_count"
  ) %>%
  left_join(meta_rural[c("SampleName", "SampleID")], by = "SampleID") %>%
  mutate(SampleName_microbe = paste(SampleName, microbe, sep = ";")) %>%
  left_join(DA_microbe_abundances2_melt[c("SampleName_microbe", "microbe_abundance_CLR")], by="SampleName_microbe") %>%
  dplyr::select(c("microbe_gene", "microbe_abundance_CLR", "residual_gene_count"))
```


```{r}
library(shiny)

# Assuming you have already loaded your data into 'residuals_melt'

# Define the UI
ui <- fluidPage(
  titlePanel("Batch + Microbe"),
  mainPanel(
    plotOutput("dotplot"),
    fluidRow(
      column(4, selectInput("gene_selector", "Select microbe_gene:", choices = unique(residuals_melt_urban$microbe_gene))), #CHANGE HERE
      column(4, actionButton("prev_plot", "Previous Plot")),
      column(4, actionButton("next_plot", "Next Plot"))
    )
  )
)

# Define the server
server <- function(input, output, session) {
  # Create a list of unique microbe_gene values
  unique_genes <- unique(residuals_melt_urban$microbe_gene) # AND HERE
  
  # Initialize a reactive value to keep track of the current index
  current_gene_index <- reactiveVal(1)
  
  # Create a reactive data subset based on the current index
  selected_data <- reactive({
    dplyr::filter(residuals_melt_urban, microbe_gene == unique_genes[current_gene_index()]) #AND HERE
  })
  
  # Create the dotplot
  output$dotplot <- renderPlot({
    ggplot(selected_data(), aes(x = microbe_abundance_CLR, y = residual_gene_count)) +
      geom_point() +
      geom_smooth(method = "lm") +
      facet_wrap(~microbe_gene, scales = "free") +
      ggtitle(paste("Dotplot for", unique_genes[current_gene_index()]))
  })
  
  # Define an observer for the "Next Plot" button
  observeEvent(input$next_plot, {
    # Increment the index to display the next plot
    current_gene_index(current_gene_index() + 1)
    
    # Wrap back to the first plot if all plots have been shown
    if (current_gene_index() > length(unique_genes)) {
      current_gene_index(1)
    }
  })
  
  # Define an observer for the "Previous Plot" button
  observeEvent(input$prev_plot, {
    # Decrement the index to display the previous plot
    current_gene_index(current_gene_index() - 1)
    
    # Wrap to the last plot if going back from the first plot
    if (current_gene_index() < 1) {
      current_gene_index(length(unique_genes))
    }
  })
  
  # Update the dropdown menu based on the current index
  observe({
    updateSelectInput(session, "gene_selector", selected = unique_genes[current_gene_index()])
  })
}

# Run the Shiny app
shinyApp(ui, server)

```














```{r}
melted_gssls_urban <- sig_list_summary_urban %>%
  #dplyr::filter(n_genes < 50) %>%
  pivot_longer(cols = c("n_U", "n_D"),
                          names_to = "l2fc_dir",
                          values_to = "n_l2fc_dir") %>%
  mutate(color = case_when(DA_urbanism == "rural" ~ "#6FBD8B", 
                           DA_urbanism == "urban" ~ "#7B2BC1")) %>%
  # First sort by number of host genes, then sort by urbanism to get colors right
  dplyr::arrange(DA_urbanism, n_genes, microbe)
melted_gssls_urban$microbe <- factor(melted_gssls_urban$microbe, levels = unique(melted_gssls_urban$microbe))

ggplot(melted_gssls_urban, aes(x = l2fc_dir, y = microbe)) + 
 geom_point(aes(size = ifelse(n_l2fc_dir == 0, NA, n_l2fc_dir)), shape=18) +
  scale_size_continuous(limits = c(0, 1000)) +
  labs(size = "Number of host genes") + 
  labs(x = "Direction of\nDGE in host", fill = "# significant\nDE host genes") +
  #facet_grid(DA_urbanism ~ ., scales="free", space="free") +
  theme_minimal() +
  theme(strip.background = element_blank(),
        strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", color=melted_gssls_urban$color),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        panel.spacing = unit(0.2, "cm")) 

#ggsave("results/figs/RNAseq-metaG-microbe/microbe_diffabun_heatmap.png", width=4.2, height=6)
```
```{r}
melted_gssls_rural <- sig_list_summary_rural %>%
  #dplyr::filter(n_genes < 50) %>%
  pivot_longer(cols = c("n_U", "n_D"),
                          names_to = "l2fc_dir",
                          values_to = "n_l2fc_dir") %>%
  mutate(color = case_when(DA_urbanism == "rural" ~ "#6FBD8B", 
                           DA_urbanism == "urban" ~ "#7B2BC1")) %>%
  # First sort by number of host genes, then sort by urbanism to get colors right
  dplyr::arrange(DA_urbanism, n_genes, microbe)
melted_gssls_rural$microbe <- factor(melted_gssls_rural$microbe, levels = unique(melted_gssls_rural$microbe))

ggplot(melted_gssls_rural, aes(x = l2fc_dir, y = microbe)) + 
 geom_point(aes(size = ifelse(n_l2fc_dir == 0, NA, n_l2fc_dir)), shape=18) +
  scale_size_continuous(limits = c(0, 5000)) +
  labs(size = "Number of host genes") + 
  labs(x = "Direction of\nDGE in host", fill = "# significant\nDE host genes") +
  #facet_grid(DA_urbanism ~ ., scales="free", space="free") +
  theme_minimal() +
  theme(strip.background = element_blank(),
        strip.text = element_blank(),
        axis.text.y = element_text(face = "italic", color=melted_gssls_rural$color),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        panel.spacing = unit(0.2, "cm")) 

#ggsave("results/figs/RNAseq-metaG-microbe/microbe_diffabun_heatmap.png", width=4.2, height=6)
```

Large heatmap
Filter for genes identified as unique 
```{r}
x <- res_list_urban
heatmap_list_urban <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], ensembl_gene_id %in% unique(sig_list_df_urban$ensembl_gene_id)) %>%
        mutate(microbe = names(x)[i])
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

heatmap_list_urban <- heatmap_list_urban[names(heatmap_list_urban) %in% names(sig_list_filt_urban)]

heatmap_list_urban <- bind_rows(heatmap_list_urban) %>%
  mutate(l2fc_sign = case_when(log2FoldChange > 0 ~ "up",
                               log2FoldChange < 0 ~ "down")) %>%
  mutate(l2fc_abs = abs(log2FoldChange)) %>%
  mutate(log2_l2fc_abs = log2(l2fc_abs)) %>%
  mutate(log2_l2fc = case_when(l2fc_sign == "up" ~ log2_l2fc_abs,
                               l2fc_sign == "down" ~ log2_l2fc_abs * -1))
```
```{r}
x <- res_list_rural
heatmap_list_rural <- foreach(
  i=seq_along(x),
  #.packages='DESeq2',
  .final = function(i) setNames(i, names(x))) %do% {
    tryCatch({
      y <-  dplyr::filter(x[[i]], ensembl_gene_id %in% unique(sig_list_df_rural$ensembl_gene_id)) %>%
        mutate(microbe = names(x)[i])
      #%>% annotate_ensembl()
      },
      error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
  }

heatmap_list_rural <- heatmap_list_rural[names(heatmap_list_rural) %in% names(sig_list_filt_rural)]

heatmap_list_rural <- bind_rows(heatmap_list_rural) %>%
  mutate(l2fc_sign = case_when(log2FoldChange > 0 ~ "up",
                               log2FoldChange < 0 ~ "down")) %>%
  mutate(l2fc_abs = abs(log2FoldChange)) %>%
  mutate(log2_l2fc_abs = log2(l2fc_abs)) %>%
  mutate(log2_l2fc = case_when(l2fc_sign == "up" ~ log2_l2fc_abs,
                               l2fc_sign == "down" ~ log2_l2fc_abs * -1)) 
```

```{r}
heatmap_plot_urban <- heatmap_list_urban %>%
  dplyr::select(ensembl_gene_id, microbe, log2FoldChange) %>%
  pivot_wider(names_from = microbe, values_from = log2FoldChange) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()

col_colors_urban <- ifelse(
  colnames(heatmap_plot_urban) %in% 
    aldex2_clr_all_sig$microbe[aldex2_clr_all_sig$DA_urbanism == "rural"],
  "#6FBD8B",
  "#7B2BC1"
)


# Update the heatmap function with cluster information
heatmap(heatmap_plot_urban,
        ColSideColors = col_colors_urban,
        col = rainbow(100))

legend(x="bottomright", legend=c("min", "ave", "max"), 
     fill=rainbow(3))
```
```{r}
library(ggfortify)

heatmap_plot_urban_data <- heatmap_plot_urban %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("microbe") %>%
  left_join(aldex2_clr_all_sig[,c("microbe", "DA_urbanism", "effect")], by="microbe") %>%
  column_to_rownames("microbe")

pca_res_urban <- prcomp(t(heatmap_plot_urban), scale. = TRUE)

autoplot(pca_res_urban, data = heatmap_plot_urban_data, colour = 'DA_urbanism')
autoplot(pca_res_urban, data = heatmap_plot_urban_data, colour = 'effect')
```

```{r}
heatmap_plot_rural <- heatmap_list_rural %>%
  dplyr::select(ensembl_gene_id, microbe, log2FoldChange) %>%
  pivot_wider(names_from = microbe, values_from = log2FoldChange) %>%
  column_to_rownames("ensembl_gene_id") %>%
  as.matrix()

col_colors_rural <- ifelse(
  colnames(heatmap_plot_rural) %in% 
    aldex2_clr_all_sig$microbe[aldex2_clr_all_sig$DA_urbanism == "rural"],
  "#6FBD8B",
  "#7B2BC1"
)


# Update the heatmap function with cluster information
heatmap(heatmap_plot_rural,
        ColSideColors = col_colors_rural,
        col = rainbow(100))

legend(x="bottomright", legend=c("min", "ave", "max"), 
     fill=rainbow(3))
```
```{r}
heatmap_plot_rural_data <- heatmap_plot_rural %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("microbe") %>%
  left_join(aldex2_clr_all_sig[,c("microbe", "DA_urbanism", "effect")], by="microbe") %>%
  column_to_rownames("microbe")

pca_res_rural <- prcomp(t(heatmap_plot_rural), scale. = TRUE)

autoplot(pca_res_rural, data = heatmap_plot_rural_data, colour = 'DA_urbanism')
autoplot(pca_res_rural, data = heatmap_plot_rural_data, colour = 'effect')
```


```{r}
library(ComplexHeatmap)
library(colorRamp2)

set.seed(123)
hm_2_urban <- ComplexHeatmap::Heatmap(heatmap_plot_urban, row_km = 2)

colname_hm_2_urban = unlist(hm_2_urban@column_names_param["labels"])

colcolor_hm_2_urban <- ifelse(colname_hm_2_urban %in% 
    aldex2_clr_all_sig$microbe[aldex2_clr_all_sig$DA_urbanism == "rural"],
  "Rural",
  "Urban"
)

anno_hm_2_urban <- HeatmapAnnotation(Urbanism = colcolor_hm_2_urban, 
                               col = list(colcolor_hm_2_urban = c("Rural" = "#6FBD8B", "Urban" = "#7B2BC1")))

hm_2_urban@row_names_param["show"] <- FALSE
hm_2_urban@column_names_param["show"] <- FALSE
hm_2_urban@top_annotation <- anno_hm_2_urban
hm_2_urban
```
```{r}
set.seed(123)
hm_2_rural <- ComplexHeatmap::Heatmap(heatmap_plot_rural, row_km = 2)

colname_hm_2_rural = unlist(hm_2_rural@column_names_param["labels"])

colcolor_hm_2_rural <- ifelse(colname_hm_2_rural %in% 
    aldex2_clr_all_sig$microbe[aldex2_clr_all_sig$DA_urbanism == "rural"],
  "Rural",
  "Urban"
)

anno_hm_2_rural <- HeatmapAnnotation(Urbanism = colcolor_hm_2_rural, 
                               col = list(colcolor_hm_2_rural = c("Rural" = "#6FBD8B", "Urban" = "#7B2BC1")))

setwd("/Users/sabri/Documents/GitHub/gmbc-colonocytes")

hm_2_rural@row_names_param["show"] <- FALSE
hm_2_rural@column_names_param["show"] <- FALSE
hm_2_rural@top_annotation <- anno_hm_2_urban
hm_2_rural
```



Functional analysis: select all genes
```{r}
library(clipr)
write_clip(gene_list_summary$hgnc_symbol)
```


Functional analysis: select genes grouped by k-means
```{r}
replace_with_row_names <- function(indices, matrix) {
  row_names <- rownames(matrix)
  row_names[indices]
}

k2_index <- row_order(hm_2)
k2_gene <- lapply(k2_index, replace_with_row_names, matrix = heatmap_plot)

k3_index <- row_order(hm_3)
k3_gene <- lapply(k3_index, replace_with_row_names, matrix = heatmap_plot)
```


```{r}
library(biomaRt)
library(clusterProfiler)

# Create a biomart object for Ensembl and UniProt databases
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Function to convert Ensembl IDs to UniProt IDs
convert_ensembl_to_symbol <- function(ensembl_ids) {
  ensembl_to_uniprot <- getBM(filters = "ensembl_gene_id",
                             values = ensembl_ids,
                             mart = ensembl,
                             attributes = c("ensembl_gene_id", "uniprot_gn_symbol", "uniprot_gn_id", "entrezgene_id"))
  ensembl_to_uniprot
}

# Apply the function to the nested list
k2_symbol <- lapply(k2_gene, convert_ensembl_to_symbol)

uni <- res_list[[1]]$ensembl_gene_id
uni <- convert_ensembl_to_symbol(uni)
```

```{r}
uni_uniprot <- unlist(list(uni$uniprot_gn_id))

# k=2 

kegg_k2_1 <- enrichKEGG(gene  = unlist(list(k2_symbol[[1]]$uniprot_gn_id)),
                      universe = uni_uniprot,
                      keyType = "uniprot",
                      organism     = 'hsa',
                      pvalueCutoff = 0.1)
kegg_k2_1_res <- kegg_k2_1@result
kegg_k2_1_sig <- kegg_k2_1_res %>% dplyr::filter(p.adjust < 0.1)
```

```{r}
cnetplot(kegg_k2_1, categorySize="p.adjust", foldChange=unlist(list(k2_symbol[[1]]$uniprot_gn_symbol)))
#ggsave("Downloads/kegg_cnet.png", width=10, height=10)
```
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
go_k2_1 <- enrichGO(gene  = unlist(list(k2_symbol[[1]]$ensembl_gene_id)),
                    universe = res_list[[1]]$ensembl_gene_id,
                    OrgDb  = org.Hs.eg.db,
                    keyType       = 'ENSEMBL',
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

go_k2_1_res <- go_k2_1@result
go_k2_1_res <- go_k2_1_res %>%
  mutate(GeneRatio_Split = strsplit(GeneRatio, "/")) %>%
  mutate(GeneRatio_Numerator = sapply(GeneRatio_Split, function(x) as.numeric(x[1]))) %>%
  mutate(GeneRatio_Denominator = sapply(GeneRatio_Split, function(x) as.numeric(x[2]))) %>%
  mutate(GeneRatio_Percentage = GeneRatio_Numerator / GeneRatio_Denominator) %>%
  mutate(BgRatio_Split = strsplit(BgRatio, "/")) %>%
  mutate(BgRatio_Numerator = sapply(BgRatio_Split, function(x) as.numeric(x[1]))) %>%
  mutate(BgRatio_Denominator = sapply(BgRatio_Split, function(x) as.numeric(x[2]))) %>%
  mutate(BgRatio_Percentage = BgRatio_Numerator / BgRatio_Denominator) %>%
  select(-GeneRatio_Numerator, -GeneRatio_Denominator, -GeneRatio_Split, -BgRatio_Numerator, -BgRatio_Denominator, -BgRatio_Split)


go_k2_1_sig <- go_k2_1_res %>% dplyr::filter(p.adjust < 0.1)
```

```{r}
goplot(go_k2_1)
```
```{r}
cnetplot(go_k2_1, categorySize="p.adjust", foldChange=unlist(list(k2_symbol[[1]]$uniprot_gn_symbol)))
#ggsave("Downloads/go_cnet.png", width=10, height=10)
```

```{r}
dotplot(go_k2_1)
```





```{r}
library(ReactomePA)

uni_entrez <- as.character(unlist(list(uni$entrezgene_id)))

react_k2_1 <- enrichPathway(unlist(list(k2_symbol[[1]]$entrezgene_id)),
                            universe=uni_entrez,
                pvalueCutoff = 0.05)
react_k2_1_res <- react_k2_1@result
react_k2_1_sig <- react_k2_1_res %>% dplyr::filter(p.adjust < 0.1)
```
```{r}
dotplot(kegg_k2_1, showCategory=15)
dotplot(react_k2_1, showCategory=15)
```
```{r}
cnetplot(react_k2_1, categorySize="p.adjust", foldChange=unlist(list(k2_symbol[[1]]$uniprot_gn_symbol)))
#ggsave("Downloads/react_cnet.png", width=10, height=10)
```

```{r}
library(enrichR)
websiteLive <- getOption("enrichR.live")
if (websiteLive) {
    listEnrichrSites()
    setEnrichrSite("Enrichr") # Human genes   
}
if (websiteLive) dbs <- listEnrichrDbs()
if (websiteLive) head(dbs)
```
```{r}
dbs <- c("MSigDB_Hallmark_2020", "GO_Biological_Process_2023", "WikiPathway_2021_Human", "BioCarta_2016", "Panther_2016")
if (websiteLive) {
    enriched <- enrichr(unlist(list(k2_symbol[[1]]$uniprot_gn_symbol)), dbs)
}
```
```{r}
if (websiteLive) msigdb_k2_1 <- enriched[["MSigDB_Hallmark_2020"]]
if (websiteLive) go23_k2_1 <- enriched[["GO_Biological_Process_2023"]]
if (websiteLive) wiki_k2_1 <- enriched[["WikiPathway_2021_Human"]]
if (websiteLive) biocarta_k2_1 <- enriched[["BioCarta_2016"]]
if (websiteLive) panther_k2_1 <- enriched[["Panther_2016"]]

plotEnrich(msigdb_k2_1, showTerms = 30, numChar = 40, y = "Count", orderBy = "Adjusted.P.value")
plotEnrich(go23_k2_1, showTerms = 30, numChar = 40, y = "Count", orderBy = "Adjusted.P.value")
plotEnrich(wiki_k2_1, showTerms = 30, numChar = 40, y = "Count", orderBy = "Adjusted.P.value")
plotEnrich(biocarta_k2_1, showTerms = 30, numChar = 40, y = "Count", orderBy = "Adjusted.P.value")
plotEnrich(panther_k2_1, showTerms = 30, numChar = 40, y = "Count", orderBy = "Adjusted.P.value")
```




```{r}
rm(list = ls())
```
